---
title: "Darwin Core mapping"
subtitle: "For: National checklists and red lists for European butterflies"
author:
- Peter Desmet
- Dimitri Brosens
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/inbo/eurobutt-checklist/blob/master/src/dwc_mapping.Rmd).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
```

# Read source data

The source data is maintained in [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/1RvxpOYf2ZrTu9nsTLumoi-G-GGhh6_lV37TNtPiVES4/edit?usp=sharing).

Read the relevant worksheets (published as csv):

```{r}
input_distribution <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSrKN8_XbQ4Vo-VrtUiqAtS5im3QxkHBNSJHPSLmrkM5C1PIC7DOg-oboRcEJZtWp_qsi802YRlRp8C/pub?gid=979140000&single=true&output=csv")
input_taxa <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSrKN8_XbQ4Vo-VrtUiqAtS5im3QxkHBNSJHPSLmrkM5C1PIC7DOg-oboRcEJZtWp_qsi802YRlRp8C/pub?gid=1559651428&single=true&output=csv")
input_regions <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSrKN8_XbQ4Vo-VrtUiqAtS5im3QxkHBNSJHPSLmrkM5C1PIC7DOg-oboRcEJZtWp_qsi802YRlRp8C/pub?gid=2076261682&single=true&output=csv")
input_references <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSrKN8_XbQ4Vo-VrtUiqAtS5im3QxkHBNSJHPSLmrkM5C1PIC7DOg-oboRcEJZtWp_qsi802YRlRp8C/pub?gid=1083424499&single=true&output=csv")
```

Sort the source files (to maintain some consistency between updates of the dataset):

```{r}
input_distribution %<>% arrange(scientific_name, region_code)
input_taxa %<>% arrange(scientific_name)
input_regions %<>% arrange(region_code)
input_references %<>% arrange(region_code, citation)
```

Copy the source data to the repository to keep track of changes:

```{r}
write_csv(input_distribution, here("data", "raw", "distribution.csv"), na = "")
write_csv(input_taxa, here("data", "raw", "taxa.csv"), na = "")
write_csv(input_regions, here("data", "raw", "regions.csv"), na = "")
write_csv(input_references, here("data", "raw", "references.csv"), na = "")
```

The first 4 files will be used for the Darwin Core mapping, while `rlc` is needed for the calculation in `wrl_values.Rmd`.

# Preprocessing

## Join spreadsheets

The `input_references` contain references of species/red lists per region. We can include those as `source` for distributions, but unfortunately not as full references in a [reference extension](http://rs.gbif.org/extension/gbif/1.0/references.xml), as that would require them to be associated (and repeated) with taxa.

Group `references` to create a concatenated list of unique references per `region_code`, e.g. `AD` = `Đug (2013) | Koren and Kulijer (2016) | Lelo (2016)`. References that act as both red and species list reference, will only be listed once:

```{r}
grouped_references <-
  input_references %>%
  arrange(region_code, citation_type, citation) %>%
  group_by(region_code) %>%
  summarize(reference = paste(unique(citation), collapse = " | "))
```

Start from distributions and join with taxon, region and grouped references information (which should result in the same number of rows as distributions):

```{r}
input_data <-
  input_distribution %>%
  left_join(input_taxa, on = "scientific_name") %>%
  left_join(input_regions, on = "region_code") %>%
  left_join(grouped_references, on = "region_code")
```

```{r}
nrow(input_data) == nrow(input_distribution)
```

## Tidy data

Clean data somewhat:

```{r}
input_data %<>%
  remove_empty("rows") %>%    # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

Sort on scientific name and region code (to maintain some consistency between updates of the dataset):

```{r}
input_data %<>% arrange(scientific_name, region_code)
```

## Scientific names

Information regarding scientific names, authorship, classification and phylogenetic order in `input_taxa` is derived from [Wiemers et al. (2018)](https://doi.org/10.3897/zookeys.811.28712), which is an "updated checklist of the European Butterflies" with 496 species.

All names on the regional checklists (`scientific_name_regional` in `input_distribution`) were matched by Dimitri Brosens to their accepted name (`scientific_name`) using the [GBIF Backbone Taxonomy](https://doi.org/10.15468/39omei) and verified by Dirk Maes (e.g. `Lycaeides argyrognomon` to official `Plebejus argyrognomon`). A limited number of `scientific_name_regional` could not be matched and/or are not accepted by Wiemers et al. (2018):

```{r}
input_data %>%
  filter(is.na(family)) %>% # Is a field from input_taxa and thus assesses there was a match
  select(scientific_name_regional, scientific_name, region_code, comments) %>%
  arrange(scientific_name_regional)
```

We **restrict the checklist to scientific names included in Wiemers et al. (2018)** (i.e. those included in `input_taxa`, see [issue 16](https://github.com/inbo/eurobutt-checklist/issues/16)):

```{r}
input_data %<>% filter(!is.na(family)) # Remove distributions/names without match
```

## Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name (that will remain the same as long as scientific name remains the same):

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "eurobutt-checklist",
  "taxon",
  vdigest(paste(scientific_name), algo = "md5"),
  sep = ":"
))
```

## Preview data

Show the number of species and distributions:

```{r}
input_data %>%
  group_by(family) %>%
  summarize(
    `# taxa` = n_distinct(taxon_id),
    `# distributions` = n()
  ) %>%
  adorn_totals("row")
```

Preview data:

```{r}
input_data %>% head()
```

# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (retaining the first of multiple distribution rows):

```{r start_taxon}
taxon <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml):
 
### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "INBO")
```

### accessRights

```{r}
taxon %<>% mutate(dwc_accessRights = "https://www.inbo.be/en/norms-data-use")
```

### datasetID

```{r}
taxon  %<>% mutate(dwc_datasetID = "https://doi.org/10.15468/ye7whj")
```

### institutionCode

```{r}
taxon %<>% mutate(dwc_institutionCode = "INBO")
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "National checklists and red lists for European butterflies")
```

### taxonID

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = scientific_name)
```

### kingdom

```{r}
taxon %<>% mutate(dwc_kingdom = "Animalia")
```

### phylum

```{r}
taxon %<>% mutate(dwc_phylum = "Arthropoda")
```

### class

```{r}
taxon %<>% mutate(dwc_class = "Insecta")
```

### order

```{r}
taxon %<>% mutate(dwc_order = "Lepidoptera")
```

### family

```{r}
taxon %<>% mutate(dwc_family = family)
```

### genus

```{r}
taxon %<>% mutate(dwc_genus = genus)
```

### specificEpithet

```{r}
taxon %<>% mutate(dwc_specificEpithet = specific_epithet)
```

### taxonRank

```{r}
taxon %<>% mutate(dwc_taxonRank = "species")
```

### scientificNameAuthorship

```{r}
taxon %<>% mutate(dwc_scientificNameAuthorship = authorship)
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(dwc_nomenclaturalCode = "ICZN")
```

## Distribution extension

Create a dataframe with all data:

```{r}
distribution <- input_data
```

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml):

### taxonID

```{r}
distribution %<>% mutate(dwc_taxonID = taxon_id) 
```

### locationID

Map values:

```{r}
distribution %<>% mutate(dwc_locationID = case_when(
  # Europe: not WGSRPD:1 as that does not include Azores and Canary islands
  # https://en.wikipedia.org/wiki/World_Geographical_Scheme_for_Recording_Plant_Distributions#1_Europe
  region_code == "EUR" ~ "",
  
  # European union: EU is reserved, but not an official ISO code
  # https://en.wikipedia.org/wiki/ISO_3166-1#Criteria_for_inclusion
  region_code == "EU27" ~ "",
  
  # Azores, Canary Islands, Madeira: use marine regions codes
  # The island groups have ISO_3166-2 subdivisions codes, but for consistency we use marine regions for all
  region_code == "MA_AZ" ~ "http://marineregions.org/mrgid/2454", # ISO_3166:PT-20
  region_code == "MA_AZ_Corvo" ~ "http://marineregions.org/mrgid/2462",
  region_code == "MA_AZ_Faial" ~ "http://marineregions.org/mrgid/2458",
  region_code == "MA_AZ_Flores" ~ "http://marineregions.org/mrgid/2461",
  region_code == "MA_AZ_Graciosa" ~ "http://marineregions.org/mrgid/2463",
  region_code == "MA_AZ_Pico" ~ "http://marineregions.org/mrgid/2460",
  region_code == "MA_AZ_Santa Maria" ~ "http://marineregions.org/mrgid/2459",
  region_code == "MA_AZ_Sao Jorge" ~ "http://marineregions.org/mrgid/2455",
  region_code == "MA_AZ_Sao Miguel" ~ "http://marineregions.org/mrgid/2456",
  region_code == "MA_AZ_Terceira" ~ "http://marineregions.org/mrgid/2457",
  
  region_code == "MA_CA" ~ "http://marineregions.org/mrgid/3743", # ISO_3166:ES-CN
  region_code == "MA_CA_El Hierro" ~ "http://marineregions.org/mrgid/3747",
  region_code == "MA_CA_Fuerteventura" ~ "http://marineregions.org/mrgid/3757",
  region_code == "MA_CA_Gran Canaria" ~ "http://marineregions.org/mrgid/3746",
  region_code == "MA_CA_La Gomera" ~ "http://marineregions.org/mrgid/3744",
  region_code == "MA_CA_La Palma" ~ "http://marineregions.org/mrgid/3748",
  region_code == "MA_CA_Lanzarote" ~ "http://marineregions.org/mrgid/3755",
  region_code == "MA_CA_Tenerife" ~ "http://marineregions.org/mrgid/3756",

  region_code == "MA_MA" ~ "http://marineregions.org/mrgid/4955", # ISO_3166:PT-30
  region_code == "MA_MA_Madeira" ~ "http://marineregions.org/mrgid/4956",
  region_code == "MA_MA_Porto Santo" ~ "http://marineregions.org/mrgid/4958",
  
  # All other countries: ISO_3166 code
  is.na(region_name) ~ paste0("ISO_3166:", region_code)
))
```

### locality

Map values:

```{r}
distribution %<>% mutate(dwc_locality = case_when(
  # Use country name if not a region  
  is.na(region_name) ~ country_name,
  
  # Use e.g. "Azores, Corvo" for island groups
  str_detect(region_code, "MA_AZ_") ~ paste("Azores", region_name, sep = ", "),
  str_detect(region_code, "MA_MA_") ~ paste("Madeira", region_name, sep = ", "),
  str_detect(region_code, "MA_CA_") ~ paste("Canary Islands", region_name, sep = ", "),
  
  # Use region name for rest
  TRUE ~ region_name
))
```

### countryCode

Map values:

```{r}
distribution %<>% mutate(dwc_countryCode = country_code)
```

Inspect mapped values: 

```{r}
distribution %>%
  group_by(
    region_code, region_name, country_code, country_name,
    dwc_locationID, dwc_locality, dwc_countryCode
  ) %>%
  count()
```

### occurrenceStatus

Inspect values:

```{r}
distribution %>%
  group_by(status) %>%
  count()
```

Map values (see [Wiemers et al. 2018](https://doi.org/10.3897/zookeys.811.28712), Materials & Methods). For migrants we decided to map to own term `migrant` as this information is too important to be lumped into `irregular` (with alternative term `vagrant`) and cannot be considered just `present` or `absent` (see  issue [#18](https://github.com/inbo/eurobutt-checklist/issues/18)).

```{r}
distribution %<>% mutate(dwc_occurrenceStatus = recode(status,
  "A" = "absent",
  "Ex" = "absent", # Regionally extinct, is indicated in threatStatus "RE"
  "Excluded" = "excluded",
  "I" = "irregular", # Irregular vagrant
  "M" = "migrant", # Regular migrant
  "P" = "present",
  "P?" = "doubtful", # Possibly present
  "P(I)" = "irregular", # Probably only present as an immigrant
  .default = ""
))
```

Inspect mapped values: 

```{r}
distribution %>%
  group_by(status, dwc_occurrenceStatus) %>%
  count()
```

### threatStatus

Inspect values:

```{r}
distribution %>%
  group_by(rlc) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_threatStatus = recode(rlc,
  # Official IUCN terms: https://www.iucnredlist.org/about/regional

  "RE" = "RE", # Regionally Extinct
  "CR" = "CR", # Critically Endangered
  "EN" = "EN", # Endangered
  "VU" = "VU", # Vulnerable
  "NT" = "NT", # Near Threatened
  "LC" = "LC", # Least Concern
  "DD" = "DD", # Data Deficient
  "NtA" = "NA", # Not Applicable
  "NE" = "NE", # Not Evaluated
  
  # Unofficial terms
  "NRLA" = "", # No red list available
  "R" = "Rare", # Used by Germany
  "Unknown" = "unknown", # not an official IUCN term, but included to be explicit
  "LC/NE" = "NE", # Used by Poland
  .default = ""
))
```

Inspect mapped values: 

```{r}
distribution %>%
  group_by(rlc, dwc_threatStatus) %>%
  count()
```

### source

```{r}
distribution %<>% mutate(dwc_source = reference)
```

### occurrenceRemarks

```{r}
distribution %<>% mutate(dwc_occurrenceRemarks = case_when(
  scientific_name_regional != scientific_name ~ paste0("In source as '", scientific_name_regional, "'")
))
```

# Vernacular names extension

Create a dataframe with unique taxa only (retaining the first of multiple distribution rows):

```{r start_vernacular_names}
vernacular_names <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Remove taxa where the vernacular name contains "suggestion":

```{r}
vernacular_names %<>% filter(!str_detect(english_name, "suggestion"))
```

Map the data to [Vernacular Names](http://rs.gbif.org/extension/gbif/1.0/vernacularname.xml):

### taxonID

```{r}
vernacular_names %<>% mutate(dwc_taxonID = taxon_id) 
```

### vernacularName

```{r}
vernacular_names %<>% mutate(dwc_vernacularName = english_name)
```

### language

```{r}
vernacular_names %<>% mutate(dwc_language = "en") 
```

# Post-processing

Only keep the Darwin Core columns:

```{r}
taxon %<>% select(starts_with("dwc_"))
distribution %<>% select(starts_with("dwc_"))
vernacular_names %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
colnames(distribution) <- str_remove(colnames(distribution), "dwc_")
colnames(vernacular_names) <- str_remove(colnames(vernacular_names), "dwc_")
```

Preview taxon core:

```{r}
taxon %>% head()
```

Preview distribution extension:

```{r}
distribution %>% head()
```

Preview vernacular names extension:

```{r}
vernacular_names %>% head()
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(vernacular_names, here("data", "processed", "vernacularname.csv"), na = "")
```
